Estimating duration using survival analysis

### 2023-05-26 cis-perfect-testing.tex
Rearrange chapter

\title{Estimating duration using survival analysis}
\author{Joshua Blake}
\date{\today}

\begin{document}
\maketitle

\pagebreak

@@

This chapter proceeds as follows.
\begin{itemize}
\item  I introduce the motivating dataset in section~\ref{perf-test:sec:data}. This is the CIS dataset previously encountered, processed to form "episodes" of infection.
\item  In section~\ref{perf-test:sec:model}, I develop a model accounting for double interval censoring and truncation. This builds on previous work in both biomedical and bird nesting studies. The arbitrary nature of truncation in CIS, where each individual has a unique truncation pattern which is neither simple left nor right truncation, has not previously been explored in the literature. Furthermore, I explore a Bayesian framework where the amount of explicit data augmentation is limited allowing the method to scale to the large number of episodes present in CIS without requiring vast computing resources (such as a high-performance cluster) which are unavailable within the secure research service (SRS, a form of trusted research environment) where the CIS data is held.
\item  In section~\ref{perf-test:sec:parameters-priors} I discuss parameterisation of the survival function and the challenges of specifying a vague prior. Whether a prior is vague or not depends on the parameterisation chosen, and for independent priors on the parameters, it is not possible to be vague on all sensible parameterisations. I also discuss the appeal of informative priors, and how the estimates in the previous chapter could be incorporated.
\item  Section~\ref{perf-test:sec:simulation-study} is a simulation study showing the applicability of this method to the CIS. I show that the model developed previously is able to estimate the duration distribution in this setting under vague or informative priors. However, I also show that, within the simulation considered, only the survival, and not the hazard, function are identifiable.
\item  Section~\ref{perf-test:sec:conclusion} concludes the chapter, discussing the limitations of this model. The following chapter will motivate and consider relaxing the perfect testing assumption.
\end{itemize}

@@

I filter the episodes as follows.
\begin{itemize}
\item
  The first positive in the episode occurred between 16th Oct 2020 and
  5th Dec 2020 inclusive.
\item
  The individual had recorded a negative test prior to their episode
  beginning (this could have happened at any time, including prior to
  16th Oct).
\item
  A negative test to end the episode (ie: following the final positive in the episode) has been recorded.
\end{itemize}

@@

\section{Modelling the duration}\label{perf-test:sec:model}
Now I develop a model accounting for double interval censoring and truncation, building on prior work in both biomedical and bird nesting studies.
The arbitrary nature of truncation in CIS, where each individual has a unique truncation pattern and is neither simple left nor right truncation, has not previously been explored in the literature.
Furthermore, I explore a Bayesian framework where the amount of explicit data augmentation is limited allowing the method to scale to the large number of episodes present in CIS without requiring vast computing resources (such as a high-performance cluster).
Only limited resources are unavailable within the secure research service (SRS, a form of trusted research environment) where the CIS data is held.
Therefore, unlike most Bayesian methods (eg: \textcite{heBayesiana}, \textcite{heBayesian}, and \textcite{caoModeling}) we avoid augmenting the data with unobserved times of infection or any information on the truncated infections; this is due to the large number of infections (>4000 in the time period considered) making this approach computationally expensive.

@@

Formally (and following \textcite{heiseyModelling}), we
define three sets which partition the space of possible infection and
recovery (ie: are strict subsets of $T \times T$) given the
observations associated with episode $i$ (graphically shown in Figure
\ref{perf-test:fig:partitionSpace}). Each episode, whether observed or not, must
fall into one of these three classes with a probability that is a
function of $\theta$ (see section \ref{perf-test:sec:likelihood}).

@@

\begin{figure}
\includegraphics[width=\textwidth]{cis-perfect-testing/images/likelihood_areas.jpg}
\caption{Admissible, $\alpha_i$, inadmissible, $\beta_i$ and
untruncated, $\Omega_i^C$ regions for an episode $i$ in an
individual whose first test was at time 0 and negative, with subsequent
negative tests at times 7, 14, and 56, and positive tests at times 21
and 28. Purple region is $\alpha_i$, orange region (which includes all
negative infection times) is $\Omega_i^C$ and unshaded region is
$\beta_i$.\label{perf-test:fig:partitionSpace} \todo[inline]{Digitalise figure of regions.}}
\end{figure}

@@

The prior $p(n_i) \propto 1/n_i$ means that the model coincides with
the frequentist likelihood-based approach, up to the priors on other
parameters~\cites[section 4.2]{dempsterMaximum}{heiseyModelling}[section 8.7.5]{gelmanBayesian}, and is
the reference prior for this quantity\autocite{heBayesiana}. Under this prior,
$\sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i) \propto \sum_{n_i=1}^\infty p_{it}^{n_i-1} = 1/(1-p_{it})$.
% (\protect\hyperlink{ref-dempsterMaximum}{Dempster, Laird, and
% Rubin} (\protect\hyperlink{ref-dempsterMaximum}{1977}) section 4.2;
% \protect\hyperlink{ref-heiseyModelling}{Heisey and Nordheim}
% (\protect\hyperlink{ref-heiseyModelling}{1995});
% \protect\hyperlink{ref-gelmanBayesian}{Gelman et al.}
% (\protect\hyperlink{ref-gelmanBayesian}{2013}) section 8.7.5)

@@

The first day individual $i$ can be positive is their day of infection, $t_i^\text{onset}$, the last day is $t_i^\text{final} = t_i^\text{onset} + d_i - 1$ where $d_i$ is the duration of positivity for individual $i$.
Outside this period, all tests are negative (ie: no false positives occur). 
Any test between $t_i^\text{onset}$ and $t_i^\text{final}$ (inclusive) is positive with probability $p_{sens}$ (that is, each test is an independent Bernoulli trial).

>>>>

\begin{document}
\chapter{Survival analysis with perfect testing} \label{perf-test}

@@

@@

I filter the episodes as follows.
\begin{itemize}
\item
  The first positive in the episode occurred between 16th Oct 2020 and
  5th Dec 2020 inclusive.
\item
  The individual had recorded a negative test prior to their episode
  beginning (this could have happened at any time, including prior to
  16th Oct).
\item
  A negative test to end the episode (i.e.: following the final positive in the episode) has been recorded.
\end{itemize}

@@

\section{Modelling the duration}\label{perf-test:sec:model}
Now I develop a model accounting for double interval censoring and truncation, building on prior work in both biomedical and bird nesting studies.
The arbitrary nature of truncation in CIS, where each individual has a unique truncation pattern and is neither simple left nor right truncation, has not previously been explored in the literature.
Furthermore, I explore a Bayesian framework where the amount of explicit data augmentation is limited allowing the method to scale to the large number of episodes present in CIS without requiring vast computing resources (such as a high-performance cluster).
Only limited resources are unavailable within the secure research service (SRS, a form of trusted research environment) where the CIS data is held.
Therefore, unlike most Bayesian methods (e.g.: \textcite{heBayesiana}, \textcite{heBayesian}, and \textcite{caoModeling}) we avoid augmenting the data with unobserved times of infection or any information on the truncated infections; this is due to the large number of infections (>4000 in the time period considered) making this approach computationally expensive.

@@

Formally (and following \textcite{heiseyModelling}), we
define three sets which partition the space of possible infection and
recovery (i.e.: are strict subsets of $T \times T$) given the
observations associated with episode $i$ (graphically shown in Figure
\ref{perf-test:fig:partitionSpace}). Each episode, whether observed or not, must
fall into one of these three classes with a probability that is a
function of $\theta$ (see section \ref{perf-test:sec:likelihood}).

@@

\begin{figure}
\includegraphics[width=\textwidth]{cis-perfect-testing/images/likelihood_areas.jpg}
\todo[inline]{Digitalise figure of regions.}
\caption{Admissible, $\alpha_i$, inadmissible, $\beta_i$ and
untruncated, $\Omega_i^C$ regions for an episode $i$ in an
individual whose first test was at time 0 and negative, with subsequent
negative tests at times 7, 14, and 56, and positive tests at times 21
and 28. Purple region is $\alpha_i$, orange region (which includes all
negative infection times) is $\Omega_i^C$ and unshaded region is
$\beta_i$.\label{perf-test:fig:partitionSpace}}
\end{figure}

@@

The prior $p(n_i) \propto 1/n_i$ means that the model coincides with the frequentist conditional likelihood approach, up to the priors on other parameters~\cites[section 4.2]{dempsterMaximum}{heiseyModelling}[section 8.7.5]{gelmanBayesian}.
Furthermore, this is the reference prior for this quantity\autocite{heBayesiana}.
Under this prior,
$\sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i) \propto \sum_{n_i=1}^\infty p_{it}^{n_i-1} = 1/(1-p_{it})$.
% (\protect\hyperlink{ref-dempsterMaximum}{Dempster, Laird, and
% Rubin} (\protect\hyperlink{ref-dempsterMaximum}{1977}) section 4.2;
% \protect\hyperlink{ref-heiseyModelling}{Heisey and Nordheim}
% (\protect\hyperlink{ref-heiseyModelling}{1995});
% \protect\hyperlink{ref-gelmanBayesian}{Gelman et al.}
% (\protect\hyperlink{ref-gelmanBayesian}{2013}) section 8.7.5)

@@

The first day individual $i$ can be positive is their day of infection, $t_i^\text{onset}$, the last day is $t_i^\text{final} = t_i^\text{onset} + d_i - 1$ where $d_i$ is the duration of positivity for individual $i$.
Outside this period, all tests are negative (i.e.: no false positives occur). 
Any test between $t_i^\text{onset}$ and $t_i^\text{final}$ (inclusive) is positive with probability $p_{sens}$ (that is, each test is an independent Bernoulli trial).
